### 2019-11-20 ch3_dynbenchmark/content.tex
technical fixes
various fixes

A plethora of TI methods has been developed over the past few years and even more are being created every month (Supplementary Table~\href{https://static-content.springer.com/esm/art\%3A10.1038\%2Fs41587-019-0071-9/MediaObjects/41587\_2019\_71\_MOESM3\_ESM.xlsx}{1}). Indeed, in several repositories listing single-cell tools, such as omictools.org \cite{henry_omictoolsinformativedirectory_2014}, the ‘awesome-single-cell’ list \cite{davis_awesomesinglecell_2018} and scRNA-tools.org \cite{zappia_exploringsinglecellrnaseq_2017}, TI methods are one of the largest categories. While each method has its own unique set of characteristics in terms of underlying algorithm, required prior information and produced outputs, two of the most distinctive differences between TI methods are whether they fix the topology of the trajectory and what type(s) of graph topologies they can detect. Early TI methods typically fixed the topology algorithmically (for example, linear \cite{bendall_singlecelltrajectorydetection_2014,schlitzer_identificationcdc1cdc2committed_2015,shin_singlecellrnaseqwaterfall_2015,campbell_bayesiangaussianprocess_2015} or bifurcating trajectories \cite{haghverdi_diffusionpseudotimerobustly_2016,setty_wishboneidentifiesbifurcating_2016}) or through parameters provided by the user \cite{trapnell_dynamicsregulatorscell_2014,matsumoto_scoupprobabilisticmodel_2016}. These methods therefore mainly focus on correctly ordering the cells along the fixed topology. More recent methods also infer the topology \cite{qiu_reversedgraphembedding_2017,street_slingshotcelllineage_2018,wolf_graphabstractionreconciles_2017}, which increases the difficulty of the problem at hand, but allows the unbiased identification of both the ordering inside a branch and the topology connecting these branches.

@@

\afterpage{%
	\clearpage% flush all other floats
	\ifodd\value{page}
%	\else% uncomment this else to get odd/even instead of even/odd
	\expandafter\afterpage% put it on the next page if this one is odd
	\fi
	{%
		\begin{figure}[tbh!]
			\centering\includegraphics[height=.75\textheight]{fig/figure_2.pdf}
			\caption{
				\textbf{A characterization of the 45 methods evaluated in this study and their overall evaluation results.}
				\textbf{a}, We characterized the methods according to the wrapper type, their required priors, whether the inferred topology is constrained by the algorithm (fixed) or a parameter (param), and the types of inferable topologies. The methods are grouped vertically based on the most complex trajectory type they can infer. \textbf{b}, The overall results of the evaluation on four criteria: accuracy using a reference trajectory on real and synthetic data, scalability with increasing number of cells and features, stability across dataset subsamples and quality of the implementation. Methods that errored on more than 50$\%$ of the datasets are not included in this figure and are shown instead in Supplementary Fig.~\href{https://github.com/dynverse/dynbenchmark_results/raw/master/08-summary/results_suppfig.pdf}{2}.
			}
			\label{fig:figure_2}
		\end{figure}
		\clearpage
		\begin{figure}[tbh!]
			\centering\includegraphics[height=.75\textheight]{fig/figure_3.pdf}
			\caption{	
				\textbf{Detailed results of the four main evaluation criteria: accuracy, scalability, stability and usability.}
				\textbf{a,} The names of the methods, ordered as in Figure \ref{fig:figure_2}. \textbf{b}, Accuracy of trajectory inference methods across metrics, dataset sources and dataset trajectory types. The performance of a method is generally more stable across dataset sources, but very variable depending on the metric and trajectory type. \textbf{c}, Predicted execution times for varying numbers of cells and features (no. of cells $\times$ no. of features). Predictions were made by training a regression model after running each method on bootstrapped datasets with varying numbers of cells and features. k, thousands; m, millions; cor, correlation. \textbf{d}, Stability results by calculating the average pairwise similarity between models inferred across multiple runs of the same method. \textbf{e}, Usability scores of the tool and corresponding manuscript, grouped per category. Off-the-shelf methods were directly implemented in R and thus do not have a usability score. 
			}
			\label{fig:figure_3}
		\end{figure}
		\clearpage
	}%
}

@@

We evaluated each method on four core aspects: (1) accuracy of a prediction, given a gold or silver standard on 110 real and 229 synthetic datasets; (2) scalability with respect to the number of cells and features (for example, genes); (3) stability of the predictions after subsampling the datasets; and (4) the usability of the tool in terms of software, documentation and the manuscript. Overall, we found a large diversity across the four evaluation criteria, with only a few methods, such as PAGA, Slingshot and SCORPIUS, performing well across the board (Figure \ref{fig:figure_2}b). We will discuss each evaluation criterion in more detail (Figure \ref{fig:figure_3} and Supplementary Fig.~\href{https://github.com/dynverse/dynbenchmark_results/raw/master/08-summary/results_suppfig.pdf}{2}), after which we conclude with guidelines for method users and future perspectives for method developers.

@@

We defined several metrics to compare a prediction to a reference trajectory (Section~\ref{sec:dynb_supn1}). Based on an analysis of their robustness and conformity to a set of rules (Section~\ref{sec:dynb_supn1}), we chose four metrics each assessing a different aspect of a trajectory (Figure \ref{fig:figure_1}d): the topology (Hamming–Ipsen–Mikhailov, HIM), the quality of the assignment of cells to branches (F1branches), the cell positions (cordist) and the accuracy of the differentially expressed features along the trajectory (wcorfeatures). The data compendium consisted of both synthetic datasets, which offer the most exact reference trajectory, and real datasets, which provide the highest biological relevance. These real datasets come from a variety of single-cell technologies, organisms and dynamic processes, and contain several types of trajectory topologies (Supplementary Table~\href{https://static-content.springer.com/esm/art\%3A10.1038\%2Fs41587-019-0071-9/MediaObjects/41587\_2019\_71\_MOESM4\_ESM.xlsx}{2}). Real datasets were classified as ‘gold standard’ if the reference trajectory was not extracted from the expression data itself, such as via cellular sorting or cell mixing \cite{tian_scrnaseqmixologybetter_2018}. All other real datasets were classified as ‘silver standard’. For synthetic datasets we used several data simulators, including a simulator of gene regulatory networks using a thermodynamic model of gene regulation \cite{schaffter_genenetweaversilicobenchmark_2011}. For each simulation, we used a real dataset as a reference, to match its dimensions, number of differentially expressed genes, drop-out rates and other statistical properties \cite{zappia_splattersimulationsinglecell_2017}.

@@

While early TI methods were developed at a time where profiling more than a thousand cells was exceptional, methods now have to cope with hundreds of thousands of cells, and perhaps soon with more than ten million \cite{svensson_exponentialscalingsinglecell_2018}. Moreover, the recent application of TI methods on multi-omics single-cell data also showcases the increasing demands on the number of features \cite{cao_jointprofilingchromatin_2018}. To assess the scalability, we ran each method on up- and downscaled versions of five distinct real datasets. We modeled the running time and memory usage using a Shape Constrained Additive Model \cite{pya_shapeconstrainedadditive_2015} (Figure \ref{fig:supfigure_4}a). As a control, we compared the predicted time (and memory) with the actual time (respectively memory) on all benchmarking datasets, and found that these were highly correlated overall (Spearman rank correlation >0.9, Figure \ref{fig:supfigure_5}, and moderately to highly correlated (Spearman rank correlation of 0.5–0.9) for almost every method, depending to what extent the execution of a method succeeded during the scalability experiments (Figure \ref{fig:figure_3}c and Supplementary Fig.~\href{https://github.com/dynverse/dynbenchmark_results/raw/master/08-summary/results_suppfig.pdf}{2}a).

@@

\begin{figure}[tbh!]
	\centering\includegraphics[width=0.6\linewidth]{fig/supfigure_5.pdf}
	\caption{
		\textbf{Agreement between actual values and predictions for execution times and memory usage.}
		We created a predictive model of the running time and memory usage based on a set of scaling datasets (left), and validated this model based on the similarity of the predictions and actual values on all benchmark datasets (right). Shown are the values for each method and dataset (n = 65618 for training, n = 11939 for test). Top left indicates the Pearson correlation coefficient.
	}
	\label{fig:supfigure_5}
\end{figure}

@@

We also assessed the memory requirements of each method (Supplementary Fig.~\href{https://github.com/dynverse/dynbenchmark_results/raw/master/08-summary/results_suppfig.pdf}{2}c). Most methods had reasonable memory requirements for modern workstations or computer clusters ($\leq$12 GB) with PAGA and STEMNET in particular having a low memory usage with both a high number of cells or a high number of features. Notably, the memory requirements were very high for several methods on datasets with high numbers of cells (RaceID/StemID, pCreode and MATCHER) or features (Monocle DDRTree, SLICE and MFA).

@@

While not directly related to the accuracy of the inferred trajectory, it is also important to assess the quality of the implementation and how user-friendly it is for a biological user \cite{taschuk_tensimplerules_2017}. We scored each method using a transparent checklist of important scientific and software development practices, including software packaging, documentation, automated code testing and publication into a peer-reviewed journal (Table \ref{tab:scoresheet}). It is important to note that there is a selection bias in the tools chosen for this analysis, as we did not include a substantial set of tools due to issues with installation, code availability and executability on a freely available platform (which excludes MATLAB). The reasons for not including certain tools are all discussed on our repository (https://github.com/dynverse/dynmethods/issues?q=label:unwrappable). Installation issues seem to be quite general in bioinformatics \cite{mangul_comprehensiveanalysisusability_2018} and the trajectory inference field is no exception.

@@

Based on the results of our benchmark, we propose a set of practical guidelines for method users (Figure \ref{fig:figure_5} and guidelines.dynverse.org). We postulate that, as a method’s performance is heavily dependent on the trajectory type being studied, the choice of method should currently be primarily driven by the anticipated trajectory topology in the data. For most use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple disconnected trajectories, cycles or a complex tree structure. In each of these use cases, our evaluation suggests a different set of optimal methods, as shown in Figure \ref{fig:figure_5}. Several other factors will also impact the choice of methods, such as the dimensions of the dataset and the prior information that is available. These factors and several others can all be dynamically explored in our interactive app (guidelines.dynverse.org). This app can also be used to query the results of this evaluation, such as filtering the datasets or changing the importance of the evaluation metrics for the final ranking.

@@

Critical to the broad applicability of TI methods is the standardization of the input and output interfaces of TI methods, so that users can effortlessly execute TI methods on their dataset of interest, compare different predicted trajectories and apply downstream analyses, such as finding genes important for the trajectory, network inference \cite{aibar_scenicsinglecellregulatory_2017} or finding modules of genes \cite{saelens_comprehensiveevaluationmodule_2018}. Our framework is an initial attempt at tackling this problem, and we illustrate its usefulness here by comparing the predicted trajectories of several top-performing methods on datasets containing a linear, tree, cyclic and disconnected graph topology (Figure \ref{fig:figure_6}). Using our framework, this figure can be recreated using only a couple of lines of R code (https://methods.dynverse.org). In the future, this framework could be extended to allow additional input data, such as spatial and RNA velocity information \cite{manno_rnavelocitysingle_2018}, and easier downstream analyses. In addition, further discussion within the field is required to arrive at a consensus concerning a common interface for trajectory models, which can include additional features such as uncertainty and gene importance.

@@

We gathered a list of 71 trajectory inference tools (Supplementary Table~\href{https://static-content.springer.com/esm/art\%3A10.1038\%2Fs41587-019-0071-9/MediaObjects/41587\_2019\_71\_MOESM3\_ESM.xlsx}{1}) by searching the literature for ‘trajectory inference’ and ‘pseudotemporal ordering’, and based on two existing lists found online: \href{https://github.com/seandavi/awesome-single-cell}{awesome-single-cell} \cite{davis_awesomesinglecell_2018} and \href{https://github.com/agitter/single-cell-pseudotime}{single-cell-pseudotime} \cite{gitter_singlecellrnaseqpseudotime_2018}. We welcome any contributions by creating an issue at \href{https://methods.dynverse.org}{https://methods.dynverse.org}.

@@

To make it easy to run each method in a reproducible manner, each method was wrapped within Docker and singularity containers (available at \href{https://methods.dynverse.org}{https://methods.dynverse.org}). These containers are automatically built and tested using Travis continuous integration (\href{https://travis-ci.org/dynverse}{https://travis-ci.org/dynverse}) and can be ran using both Docker and Singularity. For each method, we wrote a wrapper script based on example scripts or tutorials provided by the authors (as mentioned in the respective wrapper scripts). This script reads in the input data, runs the method and outputs the files required to construct a trajectory. We also created a script to generate an example dataset, which is used for automated testing.

@@

Depending on the output of a method, we used different strategies to convert the output to our model (Figure \ref{fig:supfigure_1}b). Special conversions are denoted by an asterisk and will be explained in more detail in the second list below.

@@

We gathered real datasets by searching for ‘single-cell’ at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (Supplementary Table~\href{https://static-content.springer.com/esm/art\%3A10.1038\%2Fs41587-019-0071-9/MediaObjects/41587\_2019\_71\_MOESM4\_ESM.xlsx}{2}). The scripts to download and process these datasets are available on our repository (\href{https://benchmark.dynverse.org/tree/master/scripts/01-datasets}{https://benchmark.dynverse.org/tree/master/scripts/01-datasets}). Whenever possible, we preferred to start from the raw counts data. These raw counts were all normalized and filtered using a common pipeline, as discussed later. Some original datasets contained more than one trajectory, in which case we split the dataset into its separate connected trajectory, but also generated several combinations of connected trajectories to include some datasets with disconnected trajectories in the evaluation. In the end, we included 110 datasets for this evaluation.

@@

@@

The \textit{isomorphic} score returns 1 if two graphs are isomorphic, and 0 if they were not. For this, we used the used the BLISS algorithm \cite{junttila_engineeringefficientcanonical_2007}, as implemented in the R *igraph* package.

@@

\begin{figure}[tbh!]
	\centering\includegraphics[width=\linewidth]{fig/snote1fig_8.pdf}
	\caption{
		\textbf{An illustration of ranking features based on their importance in a trajectory.} 
		(a) A MDS dimensionality reudction of a real dataset in which mouse embryonic fibroblasts (MEF) differentiate into Neurons and Myocytes. (b) The ranking of feature importances from high to low. The majority of features have a very low importance. (c) Some examples, which were also highlighted in b. Higher features in the ranking are clearly specific to certain parts of the trajectory, while features lower on the ranking have a more dispersed expression pattern.
	}
	\label{fig:snote1fig_8}
\end{figure}

>>>>

A plethora of TI methods has been developed over the past few years and even more are being created every month (Supplementary Table~\hyperref[tab:supptable_1]{3.1}). Indeed, in several repositories listing single-cell tools, such as omictools.org \cite{henry_omictoolsinformativedirectory_2014}, the ‘awesome-single-cell’ list \cite{davis_awesomesinglecell_2018} and scRNA-tools.org \cite{zappia_exploringsinglecellrnaseq_2017}, TI methods are one of the largest categories. While each method has its own unique set of characteristics in terms of underlying algorithm, required prior information and produced outputs, two of the most distinctive differences between TI methods are whether they fix the topology of the trajectory and what type(s) of graph topologies they can detect. Early TI methods typically fixed the topology algorithmically (for example, linear \cite{bendall_singlecelltrajectorydetection_2014,schlitzer_identificationcdc1cdc2committed_2015,shin_singlecellrnaseqwaterfall_2015,campbell_bayesiangaussianprocess_2015} or bifurcating trajectories \cite{haghverdi_diffusionpseudotimerobustly_2016,setty_wishboneidentifiesbifurcating_2016}) or through parameters provided by the user \cite{trapnell_dynamicsregulatorscell_2014,matsumoto_scoupprobabilisticmodel_2016}. These methods therefore mainly focus on correctly ordering the cells along the fixed topology. More recent methods also infer the topology \cite{qiu_reversedgraphembedding_2017,street_slingshotcelllineage_2018,wolf_graphabstractionreconciles_2017}, which increases the difficulty of the problem at hand, but allows the unbiased identification of both the ordering inside a branch and the topology connecting these branches.

@@

\afterpage{%
	\clearpage% flush all other floats
	\ifodd\value{page}
%	\else% uncomment this else to get odd/even instead of even/odd
	\expandafter\afterpage% put it on the next page if this one is odd
	\fi
	{%
		\begin{figure}[tbh!]
			\centering\includegraphics[height=.75\textheight]{fig/figure_2.pdf}
			\caption{
				\textbf{A characterization of the 45 methods evaluated in this study and their overall evaluation results.}
				\textbf{a}, We characterized the methods according to the wrapper type, their required priors, whether the inferred topology is constrained by the algorithm (fixed) or a parameter (param), and the types of inferable topologies. The methods are grouped vertically based on the most complex trajectory type they can infer. \textbf{b}, The overall results of the evaluation on four criteria: accuracy using a reference trajectory on real and synthetic data, scalability with increasing number of cells and features, stability across dataset subsamples and quality of the implementation. Methods that errored on more than 50$\%$ of the datasets are not included in this figure and are shown instead in Supplementary Figure~\hyperref[fig:suppfig_2]{3.1}.
			}
			\label{fig:figure_2}
		\end{figure}
		\clearpage
		\begin{figure}[tbh!]
			\centering\includegraphics[height=.75\textheight]{fig/figure_3.pdf}
			\caption{	
				\textbf{Detailed results of the four main evaluation criteria: accuracy, scalability, stability and usability.}
				\textbf{a,} The names of the methods, ordered as in Figure \ref{fig:figure_2}. \textbf{b}, Accuracy of trajectory inference methods across metrics, dataset sources and dataset trajectory types. The performance of a method is generally more stable across dataset sources, but very variable depending on the metric and trajectory type. \textbf{c}, Predicted execution times for varying numbers of cells and features (no. of cells $\times$ no. of features). Predictions were made by training a regression model after running each method on bootstrapped datasets with varying numbers of cells and features. k, thousands; m, millions; cor, correlation. \textbf{d}, Stability results by calculating the average pairwise similarity between models inferred across multiple runs of the same method. \textbf{e}, Usability scores of the tool and corresponding manuscript, grouped per category. Off-the-shelf methods were directly implemented in R and thus do not have a usability score. 
			}
			\label{fig:figure_3}
		\end{figure}
		\clearpage
	}%
}

@@

We evaluated each method on four core aspects: (1) accuracy of a prediction, given a gold or silver standard on 110 real and 229 synthetic datasets; (2) scalability with respect to the number of cells and features (for example, genes); (3) stability of the predictions after subsampling the datasets; and (4) the usability of the tool in terms of software, documentation and the manuscript. Overall, we found a large diversity across the four evaluation criteria, with only a few methods, such as PAGA, Slingshot and SCORPIUS, performing well across the board (Figure \ref{fig:figure_2}b). We will discuss each evaluation criterion in more detail (Figure \ref{fig:figure_3} and Supplementary Figure~\hyperref[fig:suppfig_2]{3.1}), after which we conclude with guidelines for method users and future perspectives for method developers.

@@

We defined several metrics to compare a prediction to a reference trajectory (Section~\ref{sec:dynb_supn1}). Based on an analysis of their robustness and conformity to a set of rules (Section~\ref{sec:dynb_supn1}), we chose four metrics each assessing a different aspect of a trajectory (Figure \ref{fig:figure_1}d): the topology (Hamming–Ipsen–Mikhailov, HIM), the quality of the assignment of cells to branches (F1branches), the cell positions (cordist) and the accuracy of the differentially expressed features along the trajectory (wcorfeatures). The data compendium consisted of both synthetic datasets, which offer the most exact reference trajectory, and real datasets, which provide the highest biological relevance. These real datasets come from a variety of single-cell technologies, organisms and dynamic processes, and contain several types of trajectory topologies (Supplementary Table~\hyperref[tab:supptable_2]{3.2}). Real datasets were classified as ‘gold standard’ if the reference trajectory was not extracted from the expression data itself, such as via cellular sorting or cell mixing \cite{tian_scrnaseqmixologybetter_2018}. All other real datasets were classified as ‘silver standard’. For synthetic datasets we used several data simulators, including a simulator of gene regulatory networks using a thermodynamic model of gene regulation \cite{schaffter_genenetweaversilicobenchmark_2011}. For each simulation, we used a real dataset as a reference, to match its dimensions, number of differentially expressed genes, drop-out rates and other statistical properties \cite{zappia_splattersimulationsinglecell_2017}.

@@

While early TI methods were developed at a time where profiling more than a thousand cells was exceptional, methods now have to cope with hundreds of thousands of cells, and perhaps soon with more than ten million \cite{svensson_exponentialscalingsinglecell_2018}. Moreover, the recent application of TI methods on multi-omics single-cell data also showcases the increasing demands on the number of features \cite{cao_jointprofilingchromatin_2018}. To assess the scalability, we ran each method on up- and downscaled versions of five distinct real datasets. We modeled the running time and memory usage using a Shape Constrained Additive Model \cite{pya_shapeconstrainedadditive_2015} (Figure \ref{fig:supfigure_4}a). As a control, we compared the predicted time (and memory) with the actual time (respectively memory) on all benchmarking datasets, and found that these were highly correlated overall (Spearman rank correlation >0.9, Figure \ref{fig:supfigure_5}, and moderately to highly correlated (Spearman rank correlation of 0.5–0.9) for almost every method, depending to what extent the execution of a method succeeded during the scalability experiments (Figure \ref{fig:figure_3}c and Supplementary Figure~\hyperref[fig:suppfig_2]{3.1}a).

@@

\begin{figure}[tbh!]
	\centering\includegraphics[width=0.6\linewidth]{fig/supfigure_5.png}
	\caption{
		\textbf{Agreement between actual values and predictions for execution times and memory usage.}
		We created a predictive model of the running time and memory usage based on a set of scaling datasets (left), and validated this model based on the similarity of the predictions and actual values on all benchmark datasets (right). Shown are the values for each method and dataset (n = 65618 for training, n = 11939 for test). Top left indicates the Pearson correlation coefficient.
	}
	\label{fig:supfigure_5}
\end{figure}

@@

We also assessed the memory requirements of each method (Supplementary Figure~\hyperref[fig:suppfig_2]{3.1}c). Most methods had reasonable memory requirements for modern workstations or computer clusters ($\leq$12 GB) with PAGA and STEMNET in particular having a low memory usage with both a high number of cells or a high number of features. Notably, the memory requirements were very high for several methods on datasets with high numbers of cells (RaceID/StemID, pCreode and MATCHER) or features (Monocle DDRTree, SLICE and MFA).

@@

While not directly related to the accuracy of the inferred trajectory, it is also important to assess the quality of the implementation and how user-friendly it is for a biological user \cite{taschuk_tensimplerules_2017}. We scored each method using a transparent checklist of important scientific and software development practices, including software packaging, documentation, automated code testing and publication into a peer-reviewed journal (Table \ref{tab:scoresheet}). It is important to note that there is a selection bias in the tools chosen for this analysis, as we did not include a substantial set of tools due to issues with installation, code availability and executability on a freely available platform (which excludes MATLAB). The reasons for not including certain tools are all discussed on our repository (\url{https://github.com/dynverse/dynmethods/issues?q=label:unwrappable}). Installation issues seem to be quite general in bioinformatics \cite{mangul_comprehensiveanalysisusability_2018} and the trajectory inference field is no exception.

@@

Based on the results of our benchmark, we propose a set of practical guidelines for method users (Figure \ref{fig:figure_5} and \url{https://guidelines.dynverse.org}). We postulate that, as a method’s performance is heavily dependent on the trajectory type being studied, the choice of method should currently be primarily driven by the anticipated trajectory topology in the data. For most use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple disconnected trajectories, cycles or a complex tree structure. In each of these use cases, our evaluation suggests a different set of optimal methods, as shown in Figure \ref{fig:figure_5}. Several other factors will also impact the choice of methods, such as the dimensions of the dataset and the prior information that is available. These factors and several others can all be dynamically explored in our interactive application (\url{https://guidelines.dynverse.org}). This application can also be used to query the results of this evaluation, such as filtering the datasets or changing the importance of the evaluation metrics for the final ranking.

@@

Critical to the broad applicability of TI methods is the standardization of the input and output interfaces of TI methods, so that users can effortlessly execute TI methods on their dataset of interest, compare different predicted trajectories and apply downstream analyses, such as finding genes important for the trajectory, network inference \cite{aibar_scenicsinglecellregulatory_2017} or finding modules of genes \cite{saelens_comprehensiveevaluationmodule_2018}. Our framework is an initial attempt at tackling this problem, and we illustrate its usefulness here by comparing the predicted trajectories of several top-performing methods on datasets containing a linear, tree, cyclic and disconnected graph topology (Figure \ref{fig:figure_6}). Using our framework, this figure can be recreated using only a couple of lines of R code (\url{https://methods.dynverse.org}). In the future, this framework could be extended to allow additional input data, such as spatial and RNA velocity information \cite{manno_rnavelocitysingle_2018}, and easier downstream analyses. In addition, further discussion within the field is required to arrive at a consensus concerning a common interface for trajectory models, which can include additional features such as uncertainty and gene importance.

@@

We gathered a list of 71 trajectory inference tools (Supplementary Table~\hyperref[tab:supptable_1]{3.1}) by searching the literature for ‘trajectory inference’ and ‘pseudotemporal ordering’, and based on two existing lists found online: \href{https://github.com/seandavi/awesome-single-cell}{awesome-single-cell} \cite{davis_awesomesinglecell_2018} and \href{https://github.com/agitter/single-cell-pseudotime}{single-cell-pseudotime} \cite{gitter_singlecellrnaseqpseudotime_2018}. We welcome any contributions by creating an issue at \href{https://methods.dynverse.org}{https://methods.dynverse.org}.

@@

To make it easy to run each method in a reproducible manner, each method was wrapped within Docker and Singularity containers (available at \href{https://methods.dynverse.org}{https://methods.dynverse.org}). These containers are automatically built and tested using Travis continuous integration (\href{https://travis-ci.org/dynverse}{https://travis-ci.org/dynverse}) and can be ran using both Docker and Singularity. For each method, we wrote a wrapper script based on example scripts or tutorials provided by the authors (as mentioned in the respective wrapper scripts). This script reads in the input data, runs the method and outputs the files required to construct a trajectory. We also created a script to generate an example dataset, which is used for automated testing.

@@

Depending on the output of a method, we used different strategies to convert the output to our model (Figure \ref{fig:supfigure_1}b). Special conversions are denoted by an asterisk (*) and will be explained in more detail in the second list below.

@@

We gathered real datasets by searching for ‘single-cell’ at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (Supplementary Table~\hyperref[tab:supptable_2]{3.2}). The scripts to download and process these datasets are available on our repository (\href{https://benchmark.dynverse.org/tree/master/scripts/01-datasets}{https://benchmark.dynverse.org/tree/master/scripts/01-datasets}). Whenever possible, we preferred to start from the raw counts data. These raw counts were all normalized and filtered using a common pipeline, as discussed later. Some original datasets contained more than one trajectory, in which case we split the dataset into its separate connected trajectory, but also generated several combinations of connected trajectories to include some datasets with disconnected trajectories in the evaluation. In the end, we included 110 datasets for this evaluation.

@@

\section{Supplementary Figures and Tables}

\paragraph{Supplementary Table 3.1: Overview of available trajectory inference tools, and whether they were included in	this study.} \label{tab:supptable_1}
Download link: \url{https://static-content.springer.com/esm/art\%3A10.1038\%2Fs41587-019-0071-9/MediaObjects/41587\_2019\_71\_MOESM3\_ESM.xlsx}.

\paragraph{Supplementary Table 3.2: Overview of the real datasets used in this study.} \label{tab:supptable_2}
Download link:
\url{https://static-content.springer.com/esm/art\%3A10.1038\%2Fs41587-019-0071-9/MediaObjects/41587\_2019\_71\_MOESM4\_ESM.xlsx}.

\paragraph{Supplementary Figure 3.1: Results from the evaluation, for all methods and across all evaluation criteria.} \label{fig:suppfig_2}
(a) We characterized the methods according to the wrapper type, their required priors, whether the inferred topology is constrained by the algorithm (fixed) or a parameter (param), and the types of inferable topologies. The methods are grouped vertically based on the most complex trajectory type the y can infer. (b) The overall results of the evaluation on four criteria: benchmarking using a reference trajectory on real and synthetic data, scalability with increasing number of cells and features, stability across dataset subsamples, and quality  of  the  implementation.  (c)  Accuracy  of  trajectory  inference  methods  across metrics,  dataset sources  and  dataset  trajectory types. The performance of a method is generally more stable across dataset sources, but very  variable depending on the metric and trajectory  type.  (d)  Predicted  execution  times  and memory  usage  for  varying  numbers  of  cells  and  features  (\# cells $\times$ \# features). Predictions were made by training a regression model after running each method on bootstrapped datasets with varying numbers of cells and features. (e) Stability results by calculating the average pairwise similarity between models inferred across multiple runs of the same method. (f) Usability scores of the tool and corresponding manuscript, grouped per category.
Download link: \url{https://github.com/dynverse/dynbenchmark_results/raw/master/08-summary/results_suppfig.pdf}.

@@

The \textit{isomorphic} score returns 1 if two graphs are isomorphic, and 0 if they were not. For this, we used the used the BLISS algorithm \cite{junttila_engineeringefficientcanonical_2007}, as implemented in the R igraph package.

@@

\begin{figure}[tbh!]
	\centering\includegraphics[width=\linewidth]{fig/snote1fig_8.pdf}
	\caption{
		\textbf{An illustration of ranking features based on their importance in a trajectory.} 
		(a) A MDS dimensionality reduction of a real dataset in which mouse embryonic fibroblasts (MEF) differentiate into Neurons and Myocytes. (b) The ranking of feature importances from high to low. The majority of features have a very low importance. (c) Some examples, which were also highlighted in b. Higher features in the ranking are clearly specific to certain parts of the trajectory, while features lower on the ranking have a more dispersed expression pattern.
	}
	\label{fig:snote1fig_8}
\end{figure}
